Rao-Blackwellised Two-filter algorithm for Conditionally Linear and Gaussian models

### 2016-10-31 TexArxiv/clgm.tex
no comment

\subsection{Forward filter}
\label{sec:forward}
The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using the standard Rao Blackwellised  algorithm given in \cite{chen:liu:2000}. The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
\[
p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
\]
In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter. 
\paragraph{Initialization}
At time $i=1$, $(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to $(\pi_j)_{1\le j\le J}$. Then, $\mu_1^k$ and $P_1^k$ are computed using a Kalman filter:
\begin{align*}
K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
\end{align*}
Particle $a_1^k$ is associated with the importance weight $\omega^k_1 \propto p(y_1|a_1^k)$. This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$. 
\paragraph{Iterations}
For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.  Labels trajectories at time $i-1$ are defined by:
\[
a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
\]
where
\[
\mathsf{A}^k_{i-1} = I^k_i \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
\]
Similarly, means and variances trajectories are defined by: 
\[
\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
\]
The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
\[
\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
\]
$\mu_i^k$ and $P_i^k$ are updated according to:
\begin{align*}
\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
\end{align*}
The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.

%The conditional distribution of $Y_1$ given $a^1_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^j$ may be written:
%\begin{align*}
%\gamma_i^j &= p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j) \eqsp, \\
%& =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp.
%\end{align*}
%\begin{algorithm}
%\DontPrintSemicolon
%\KwIn{Initial distribution $\pi$ on $\{1,\ldots,J\}$, initial mean $\mu_1$ and variance $\Sigma_1$.}
%\KwOut{A set of trajectories $(a^k_{1:i})_{1\le k\le N}$, means $(\mu^k_{i})_{1\le k\le N}$ and variances $(P^k_{i})_{1\le k\le N}$ with importance weights $(\omega^k_i)_{1\le k\le N}$, $1\le i \le n$.}
%{\bf Initialization}\;
%\For{$k \gets 1$ \textbf{to} $N$} {
%Sample $a^k_1 \in \{1,\ldots,J\}$ such that $\mathbb{P}\left(a_1^k = j\right) = \pi_j$. \;
%Compute $\mu_1^k$ and $P_1^k$:
%\begin{align*}
%K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
%\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
%P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
%\end{align*}
%Update the importance weight: $\omega^k_1 \propto p(y_1|a_1^k)$.
%}
%{\bf Iterations}\;
%\For{$i \gets 2$ \textbf{to} $n$} {
%Resample the set $(\{a^k_{1:i-1}, \mu^k_{i-1}, P^k_{i-1}\})_{1\le k\le N}$ using the normalized weights $(\omega^k_{i-1})_{1\le k \le N}$.\;
%\For{$k \gets 1$ \textbf{to} $N$} {
%For all $j \in \{1,\ldots, J\}$ compute
%\[
%\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) q(a_{i-1}^k,j)\eqsp.
%\]
%Sample $a_i^k \in \{1,\ldots, J\}$ using the normalized weights $(\gamma_i^{j,k})_{1\le j\le J}$.\;
%Update $\mu_i^k$ and $P_i^k$:
%\begin{align*}
%\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
%P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp,\\
%K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
%\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
%P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
%\end{align*}
%Update the importance weights: $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.
%}
%}

%The recursive updates of $(a^k_{1:i})_{1\le k \le N}$, $(\omega^k_{i})_{1\le k \le N}$,  $\mu^k_{i}$ and $P^k_{i}$ are  displayed in Algorithm~\ref{algo:forward}.
%In the initialization step, the first weight is given, for all $1\le k \le N$, by $\omega^k_1 = p(y_1|a^k_{1})$. By \eqref{eq:model:obs}, the conditional distribution of $Y_1$ given $a_1$ is a Gaussian distribution with mean $c_{a_1}+B_{a_1}\mu_1$ and variance $B_{a_1}\Sigma_1B'_{a_1} + \barG_{a_1}$. This yields:
%\[
%\omega^k_{1} \propto \exp\left\{-\frac{1}{2}\normMat{B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}}{Y_1-c_{a^k_1}-B_{a_1^k}\mu_1}^2\right\} \eqsp,
%\]

%By Algorithm~\ref{algo:forward}, $p^N_{}(a_{i},z_{i}|y_{1:i})$ is given by:
%\begin{equation*}
%p^N_{}(a_{i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{i}}(a_{i})\omega^k_{i}|2\pi P^k_{i}|^{-1/2}\times \exp\left\{-\frac{1}{2}\normMat{P_{i}^k}{z_{i}-\mu^k_{i}}^2\right\}\eqsp.
%\end{equation*}

\subsection{Path-space smoother of \cite{kitagawa:1996}}
As noted in \cite{kitagawa:1996}, a smoother may be obtained as a byproduct of the forward filter by keeping track of the genealogy of the weighted samples obtained at each time step. In this case the whole trajectories $(a^k_{1:n})_{1\le k\le n}$ are used to approximate the joint smoothing distributions :
\[
p^{\mathsf{Kit}}_N(a_{1:n}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^k_{1:n}}(a_{1:n})\eqsp.
\]
In this case, the marginal smoothing distribution for a given time $i$ is obtained using the genealogical indices $(\mathsf{A}^k_i)_{1\le k \le N}$:
\[
p^{\mathsf{Kit}}_N(a_{i}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^{\mathsf{A}^k_i}_{i}}(a_{i})\eqsp.
\]
This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps.

\subsection{Algorithm based on the structural approximation of \cite{kim:1994}}
Another method to approximate the smoothing distributions of the indicator variables given all the observations proposed by \cite{kim:1994} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$. Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$, approximating the filtering distributions are available. Then, the approximation of \cite{kim:1994} replaces $p(a_i|a_{i+1:n},y_{1:n})$ by
\[
p(a_i|a_{i+1},y_{1:i}) \propto Q(a_i,a_{i+1})p(a_i|y_{1:i})\eqsp.
\]
Plugging the particle approximation of $p(a_i|y_{1:i})$ in this expression provides the following procedure to sample trajectories approximatively distributed according to $p(a_{1:n}|y_{1:n})$. $\tilde{a}_{n}$ is first sampled in $\{a_n^1,\ldots,a_n^N\}$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$. Then, for all $1\le i \le n-1$, $\tilde{a}_{i}$ is sampled in $\{a_i^1,\ldots,a_i^N\}$ with probabilities proportional to $(\omega_i^kQ(a_i^k,\tilde{a}_{i+1}))_{1\le k \le N}$. If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the indicator variables is:
\[
p^{\mathsf{Kim}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
\]

\subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm. The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
\[
p(a_{1:n}|y_{1:n}) = p(a_{1:i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp,
\]
where
\[
p(a_{1:i}|a_{i+1:n},y_{1:n}) \propto p(y_{i+1:n},a_{i+1:n}|a_{1:i},y_{1:i})p(a_{1:i}|y_{1:i})\eqsp.
\]
The second factor in the last equation may be approximated using the trajectories $(a^k_{1:i})_{1\le k \le N}$ and importance weights $(\omega^k_{i})_{1\le k \le N}$ produced by the forward filter. Therefore, $p(a_{1:i}|a_{i+1:n},y_{1:n})$ may be approximated by:
\[
p^N(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \tilde{\omega}^k_{i|n}\delta_{a^k_{1:i}}(a^k_{1:i})\eqsp,
\]
with $\tilde{\omega}^k_{i|n} \propto \omega_i^k p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$. Then, a trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$ may be sampled following these steps:
\begin{enumerate}[-]
\item Set $\tilde{a}_n= a_n^k$ with probability $\omega_n^k$, $1\le k \le N$.
\item For all $1\le i\le n-1$, set $\tilde{a}_i = a_i^k$ with probability $\tilde{\omega}_i^k$, $1\le k \le N$.
\end{enumerate}
This algorithm requires to compute the quantity  $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$. As noted by \cite{lindsten:bunch:sarkka:schon:godsill:2015}, this predictive quantity is available analytically using backward Kalman filtering techniques from time $n$ to time $i$. However, this has to be done for each trajectory $a^k_{1:i}$, $1\le k \le N$, which leads to an algorithm with a prohibitive computational complexity. \cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a procedure computationally less intensive by noting that:
\begin{equation}
\label{eq:FFBS:incremental}
p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i}) = \int p(y_{i+1:n},a_{i+1:n}|z_i,a^k_{i})p(z_{i}|a^k_{1:i},y_{1:i})\rmd z_i\eqsp,
\end{equation}
which is the two-filter decomposition of the smoothing distribution. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
\[
p(y_{i+1:n},a_{i+1:n}|z_i,a_{i}) \propto Q(a_i,a_{i+1}) \exp\left\{-(z_i'\Omega_iz_i-2\lambda_i'z_i)/2\right\}\eqsp,
\]
where the proportionality is with respect to $(a_i,z_i)$ and
\[
p(y_{i:n},a_{i+1:n}|z_i,a_{i}) \propto \exp\left\{-(z_i'\hat{\Omega}_iz_i-2\hat{\lambda}_i'z_i)/2\right\}\eqsp,
\]
where the proportionality is with respect to $z_i$. 
\begin{align*}
\hat{\Omega}_n &= B'_{a_n}\barG^{-1}_{a_n}B_{a_n}\eqsp,\\
\hat{\lambda}_n &=B'_{a_n}\barG^{-1}_{a_n}(y_n-c_{a_n})\eqsp.
\end{align*}
Then, for $1\le i\le n-1$, define $m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{a_{i+1}}$ and $M_{i+1} = H_{a_{i+1}}'\hat{\Omega}_{i+1}H_{a_{i+1}} + I$. Then,
\begin{align*}
\Omega_i &= T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})\hat{\Omega}_{i+1}T_{a_{i+1}}\eqsp,\\
\lambda_i &=T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})m_{i+1}\eqsp.
\end{align*}
As $p(y_{i:n},a_{i+1:n}|z_i,a_{i}) = p(y_{i}|z_i,a_{i})p(y_{i+1:n},a_{i+1:n}|z_i,a_{i})$,
\begin{align*}
\hat{\Omega}_i &= \Omega_i + B'_{a_i}\barG^{-1}_{a_i}B_{a_i}\eqsp,\\
\hat{\lambda}_i & = \lambda_i + B'_{a_i}\barG^{-1}_{a_i}(y_i-c_{a_i})\eqsp.
\end{align*}
Then, by \eqref{eq:FFBS:incremental},
\[
p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})\propto Q(a_i^k,a_{i+1})|\Lambda_i|^{-1/2}\exp\left\{\eta_i/2\right\}\eqsp,
\]
where the proportionality is with respect to $a^k_{1:i}$ and $\Lambda_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I$ and $\eta_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}$. These backward simulation steps are summarized in Algorithm~\ref{algo:FFBS}.

@@

Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}. Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods. For $3\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:

>>>>

\subsection{Forward filter}
\label{sec:forward}
The SMC approximation $p^N_{}(a_{1:i},z_{i}|y_{1:i})$ of $p_{}(a_{1:i},z_{i}|y_{1:i})$ may be obtained using the standard Rao Blackwellised  algorithm given in \cite{chen:liu:2000}. The procedure produces a sequence of trajectories $(a^k_{1:i})_{1\le k \le N}$ associated with importance weights $(\omega^k_{i})_{1\le k \le N}$ used to define the following approximation of $p_{}(a_{1:i},z_{i}|y_{1:i})$:
\[
p^N_{}(a_{1:i},z_{i}|y_{1:i}) = \sum_{k=1}^N\delta_{a^k_{1:i}}(a_{1:i})\,\omega^k_{i}\,p_{}(z_{i}|a^k_{1:i},y_{1:i})\eqsp.
\]
In this equation, the conditional distribution of the hidden state $z_{i}$ given the observations $y_{1:i}$ and a trajectory $a^k_{1:i}$ is a Gaussian distribution whose mean $\mu^k_{i}$ and variance $P^k_{i}$ may be obtained by running a Kalman filter. 
\paragraph{Initialization}
At time $i=1$, $(a^k_1)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,J\}$ with probabilities proportional to $(\pi_j)_{1\le j\le J}$. Then, $\mu_1^k$ and $P_1^k$ are computed using a Kalman filter:
\begin{align*}
K^k_{1} &=\Sigma_1B'_{a_1^k}\left(B_{a_1^k}\Sigma_1B'_{a_1^k} + \barG_{a_1^k}\right)^{-1}\eqsp,\\
\mu^k_{1} &= \mu_{1} + K^k_1\left(Y_1 - c_{a_{1}^k} - B_{a_1^k}\mu_{1}\right)\eqsp,\\
P^k_{1} &=\left(I-K^k_{1}B_{a_1^k}\right)\Sigma_1\eqsp.
\end{align*}
Particle $a_1^k$ is associated with the importance weight $\omega^k_1 \propto p(y_1|a_1^k)$. This weight is updated by noting that the conditional distribution of $Y_1$ given $a_1^k$ is a Gaussian distribution with mean $c_{a^k_1}+B_{a^k_1}\mu_1$ and variance $B_{a^k_1}\Sigma_1B'_{a^k_1} + \barG_{a^k_1}$. 
\paragraph{Iterations}
For all $2\le i \le n$, particle indices $(I^k_i)_{1\le k \le N}$ are sampled independently in $\{1,\ldots,N\}$ with probabilities proportional to  $(\omega^k_{i-1})_{1\le k \le N}$.  Labels trajectories at time $i-1$ are defined by:
\[
a_{1:i-1}^k = (a^{\mathsf{A}^k_1}_1,\ldots,a^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp,
\]
where
\[
\mathsf{A}^k_{i-1} = I^k_i \quad\mbox{and for } 1\le j<i\quad \mathsf{A}^k_{j} = I^{\mathsf{A}^k_{j+1}}_{j+1}\eqsp.
\]
Similarly, means and variances trajectories are defined by: 
\[
\mu_{1:i-1}^k = (\mu^{\mathsf{A}^k_1}_1,\ldots,\mu^{\mathsf{A}^k_{i-1}}_{i-1})\quad\mbox{and}\quad P_{1:i-1}^k = (P^{\mathsf{A}^k_1}_1,\ldots,P^{\mathsf{A}^k_{i-1}}_{i-1})\eqsp.
\]
The new indicator variables $(a_i^k)_{1\le k \le N}$ are sampled independently in  $\{1,\ldots, J\}$ with probabilities proportional to $(\gamma_i^{j,k})_{1\le j\le J}$, where for all $j \in \{1,\ldots, J\}$:
\[
\gamma_i^{j,k} = p(y_i | a_i = j, a_{1:i-1}^k, y_{1:i-1}) Q(a_{i-1}^k,j)\eqsp.
\]
The conditional distribution of $Y_i$ given $a^k_{1:i-1}$, $a_i$ and $Y_{1:i-1}$ is a Gaussian distribution with mean $c_{a_i}+B_{a_i}\mu^k_{i|i-1}$ and variance $B_{a_i}P^k_{i|i-1}B'_{a_i} + \barG_{a_i}$ which means that $\gamma_i^{j,k}$ may be written:
\[
\gamma_i^{j,k} =  Q(a_{i-1}^k,j)|B_jP^k_{i|i-1}B'_j + \barG_j|^{-1/2}\exp\left\{-\frac{1}{2}\normMat{B_jP^k_{i|i-1}B'_j + \barG_j}{y_i-c_{j}-B_j\mu^k_{i|i-1}}^2\right\} \eqsp,
\]
where 
\begin{align*}
\mu^{k}_{i|i-1} &= d_{a_{i}^k} + T_{a_{i}^k}\mu^k_{i-1}\eqsp,\\
P^k_{i|i-1} &= T_{a_{i}^k}P^k_{i-1}T'_{a_{i}^k} + \barH_{a_{i}^k}\eqsp.
\end{align*}
$\mu_i^k$ and $P_i^k$ are updated according to:
\begin{align*}
K^k_{i} &=P^k_{i|i-1}B'_{a_{i}^k}\left(B_{a_{i}^k}P^k_{i|i-1}B'_{a_{i}^k} + \barG_{a_{i}^k}\right)^{-1}\eqsp,\\
\mu^k_{i} &= \mu^{k}_{i|i-1} + K^k_i\left(Y_i - c_{a_{i}^k} - B_{a_{i}^k}\mu^{k}_{i|i-1}\right)\eqsp,\\
P^k_{i} &=\left(I-K^k_{i}B_{a_{i}^k}\right)P^k_{i|i-1}\eqsp.
\end{align*}
The importance weights are given by $\omega_i^k \propto \sum_{j=1}^J \gamma_i^{j,k}$.

\subsection{Path-space smoother of \cite{kitagawa:1996}}
As noted in \cite{kitagawa:1996}, a smoother may be obtained as a byproduct of the forward filter by keeping track of the genealogy of the weighted samples obtained at each time step. In this case the whole trajectories $(a^k_{1:n})_{1\le k\le n}$ are used to approximate the joint smoothing distributions :
\[
p^{\mathsf{Kit}}_N(a_{1:n}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^k_{1:n}}(a_{1:n})\eqsp.
\]
In this case, the marginal smoothing distribution for a given time $i$ is obtained using the genealogical indices $(\mathsf{A}^k_i)_{1\le k \le N}$:
\[
p^{\mathsf{Kit}}_N(a_{i}|Y_{1:n}) = \sum_{k=1}^n \omega^k_{n}\delta_{a^{\mathsf{A}^k_i}_{i}}(a_{i})\eqsp.
\]
This is a  poor approximation of the marginal smoothing distributions for small values of $i$ as the depletion of the particles increases with the successive resampling steps. See Section~\ref{sec:numerical:experiments} for a numerical illustration.

\subsection{Algorithm based on the structural approximation of \cite{kim:1994}}
Another method to approximate the smoothing distributions of the indicator variables given all the observations proposed by \cite{kim:1994} relies on an tractable approximation of $p(a_i|a_{i+1:n},y_{1:n})$. Assume first that the particles and importance weights $(a_i^k,\omega_i^k)_{1\le k \le N}$, $1\le i\le n$, approximating the filtering distributions are available. Then, note that, for all $1\le i\le n$,
\[
p(a_{i:n}|y_{1:n}) = p(a_{i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp.
\]
Then, the approximation of \cite{kim:1994} replaces $p(a_i|a_{i+1:n},y_{1:n})$ by
\[
p(a_i|a_{i+1},y_{1:i}) \propto Q(a_i,a_{i+1})p(a_i|y_{1:i})\eqsp.
\]
Plugging the particle approximation of $p(a_i|y_{1:i})$ in this expression provides the following procedure to sample trajectories approximatively distributed according to $p(a_{1:n}|y_{1:n})$:
\begin{enumerate}[-]
\item $\tilde{a}_{n}$ is first sampled in $\{a_n^1,\ldots,a_n^N\}$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$. 
\item Then, for all $1\le i \le n-1$, $\tilde{a}_{i}$ is sampled in $\{a_i^1,\ldots,a_i^N\}$ with probabilities proportional to $(\omega_i^kQ(a_i^k,\tilde{a}_{i+1}))_{1\le k \le N}$. 
\end{enumerate}
If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{kim:1994} of the joint smoothing distribution of the indicator variables is:
\[
p^{\mathsf{Kim}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.
\]

\subsection{FFBS based algorithm of \cite{lindsten:bunch:sarkka:schon:godsill:2015}}
\cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a Rao-Blackwellised procedure to sample the indicator variables backward in time following the same steps as in the FFBS algorithm without using the approximation of \cite{kim:1994}. The algorithm relies on the decomposition given, for all $1\le i \le n-1$, by:
p(a_{1:n}|y_{1:n}) = p(a_{1:i}|a_{i+1:n},y_{1:n})p(a_{i+1:n}|y_{1:n})\eqsp,
p(a_{1:i}|a_{i+1:n},y_{1:n}) \propto p(y_{i+1:n},a_{i+1:n}|a_{1:i},y_{1:i})p(a_{1:i}|y_{1:i})\eqsp.
The second factor in the last equation may be approximated using the trajectories $(a^k_{1:i})_{1\le k \le N}$ and importance weights $(\omega^k_{i})_{1\le k \le N}$ produced by the forward filter. Therefore, $p(a_{1:i}|a_{i+1:n},y_{1:n})$ may be approximated by:
p^N(a_{1:i}|a_{i+1:n},y_{1:n}) = \sum_{k=1}^N \tilde{\omega}^k_{i|n}\delta_{a^k_{1:i}}(a^k_{1:i})\eqsp,
with $\tilde{\omega}^k_{i|n} \propto \omega_i^k p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$. Then, a trajectory $\tilde{a}_{1:n}$ approximatively distributed according to $p(a_{1:n}|y_{1:n})$ may be sampled following these steps:
\item Set $\tilde{a}_n= a_n^k$ with probabilities proportional to $(\omega_n^k)_{1\le k \le N}$.
\item For all $1\le i\le n-1$, set $\tilde{a}_i = a_i^k$ with probabilities proportional to $(\tilde{\omega}_i^k)_{1\le k \le N}$.
This algorithm requires to compute the quantity  $p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})$. As noted by \cite{lindsten:bunch:sarkka:schon:godsill:2015}, this predictive quantity is available analytically using backward Kalman filtering techniques. However, this has to be done for each trajectory $(a^k_{1:i})_{1\le k \le N}$, which leads to an algorithm with a prohibitive computational complexity. \cite{lindsten:bunch:sarkka:schon:godsill:2015} proposed a procedure computationally less intensive by noting that:
\label{eq:FFBS:incremental}
p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i}) = \int p(y_{i+1:n},a_{i+1:n}|z_i,a^k_{i})p(z_{i}|a^k_{1:i},y_{1:i})\rmd z_i\eqsp,
which is similar to the two-filter decomposition of the smoothing distribution. By  \cite{lindsten:bunch:sarkka:schon:godsill:2015},
p(y_{i+1:n},a_{i+1:n}|z_i,a_{i}) \propto Q(a_i,a_{i+1}) \exp\left\{-(z_i'\Omega_iz_i-2\lambda_i'z_i)/2\right\}\eqsp,
where the proportionality is with respect to $(a_i,z_i)$ and
p(y_{i:n},a_{i+1:n}|z_i,a_{i}) \propto \exp\left\{-(z_i'\hat{\Omega}_iz_i-2\hat{\lambda}_i'z_i)/2\right\}\eqsp,
where the proportionality is with respect to $z_i$. These quantities may be computed recursively backward in time with:
\hat{\Omega}_n &= B'_{a_n}\barG^{-1}_{a_n}B_{a_n}\eqsp,\\
\hat{\lambda}_n &=B'_{a_n}\barG^{-1}_{a_n}(y_n-c_{a_n})\eqsp.
Then, for $1\le i\le n-1$, define $m_{i+1} = \hat{\lambda}_{i+1} - \hat{\Omega}_{i+1}d_{a_{i+1}}$ and $M_{i+1} = H_{a_{i+1}}'\hat{\Omega}_{i+1}H_{a_{i+1}} + I$ and write
\Omega_i &= T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})\hat{\Omega}_{i+1}T_{a_{i+1}}\eqsp,\\
\lambda_i &=T'_{a_{i+1}}(I-\hat{\Omega}_{i+1}H_{a_{i+1}}M^{-1}_{i+1}H'_{a_{i+1}})m_{i+1}\eqsp.
As $p(y_{i:n},a_{i+1:n}|z_i,a_{i}) = p(y_{i}|z_i,a_{i})p(y_{i+1:n},a_{i+1:n}|z_i,a_{i})$,
\hat{\Omega}_i &= \Omega_i + B'_{a_i}\barG^{-1}_{a_i}B_{a_i}\eqsp,\\
\hat{\lambda}_i & = \lambda_i + B'_{a_i}\barG^{-1}_{a_i}(y_i-c_{a_i})\eqsp.
Then, by \eqref{eq:FFBS:incremental},
p(y_{i+1:n},a_{i+1:n}|a^k_{1:i},y_{1:i})\propto Q(a_i^k,a_{i+1})|\Lambda_i|^{-1/2}\exp\left\{\eta_i/2\right\}\eqsp,
where the proportionality is with respect to $a^k_{1:i}$ and $\Lambda_i = (\Gamma_i^k)'\Omega_i\Gamma_i^k + I$ and $\eta_i = \|\mu_i^k\|^2_{\Omega^{-1}_i} - 2\lambda'_i\mu_i^k-\|(\Gamma_i^k)'(\lambda_i-\Omega_i\mu_i^k)\|^2_{\Lambda_i}$ where $P_i^k = \Gamma_i^k(\Gamma_i^k)'$.
If $(\tilde{a}^k_{1:n})_{1\le k \le \tilde{N}}$ are independent copies of $\tilde{a}_{1:n}$, the SMC approximation of \cite{lindsten:bunch:sarkka:schon:godsill:2015} of the joint smoothing distribution of the indicator variables is:
p^{\mathsf{Lbscg}}_N(a_{1:n}|Y_{1:n}) = \frac{1}{\tilde{N}}\sum_{k=1}^{\tilde N} \delta_{\tilde{a}^{k}_{1:n}}(a_{1:n})\eqsp.

@@

Following \cite{briers:doucet:maskell:2010}, this paper proposes a generalized two-filter algorithm which exploits the specific properties of the model defined by \eqref{eq:model:state} and \eqref{eq:model:obs}. Contrary to the previous methods, two-filter based smoothers are designed to compute approximations of marginal smoothing distributions (usually the posterior distribution of one or two consecutive  indicator variables given all the observations). 
%Conditional on $(a_i)_{1\le i\le n}$, these equations define a linear Gaussian model for which analytic computations may be performed to marginalize the states $(Z_{i})_{1\le i \le n}$ in order to reduce the the variance of SMC methods. 
For $3\le i \le n-1$, our proposed algorithm relies on the explicit marginalization:
